### 2013-06-15 doc/pgas13/conclusion.tex
Fix typos and improve wording and typography (based on Preston's feedback)
rewriting to conserve space, tighten language

\section{Conclusion}
In PGAS implementations there is typically a large discrepancy between the first, simple, description of an algorithm and the final, optimized one. At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
We have shown that the additional concurrency that comes with a latency-tolerant runtime, rather than compounding the problem, provides a new opportunity for reducing communication by combining locally.
% perform some of the locality optimizations that finely-tuned implementations would.
This allows flat-combined sequentially-consistent global data structures to scale well out to a thousand cores with millions of concurrent threads.

>>>>

\section{Conclusion}
In PGAS implementations there is typically a large discrepancy between the first, simple, description of an algorithm and the final, optimized one.
% At the core of this work is the desire to have a library of data structures which can be used without worrying about when they will be consistent or whether they will limit scalability.
Bringing findings from shared memory into the PGAS domain brings with it new challenges and opportunities.
We have shown that the additional concurrency that comes with a latency-tolerant runtime, rather than compounding the problem, provides a new opportunity for reducing communication by combining locally.
% perform some of the locality optimizations that finely tuned implementations would.
This allows easy implementation of a library of flat-combined linearizable global data structures that scale well out to a thousand cores and millions of concurrent threads.

### 2013-06-15 doc/pgas13/evaluation.tex
Simple space hacking w/o sacrificing content.
rewriting to conserve space, tighten language

\section{Evaluation}
To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two graph benchmarks.

Experiments were run on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand network cards, connected via a QLogic InfiniBand switch.

@@

\paragraph{Queue and Stack}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/vector_perf.pdf}
  \caption{\emph{Queue and Stack performance.}
    Results are shown on a log scale for a throughput workload performing 256 million operations with 2048 workers per core and 16 cores per node. Local flat combining improves throughput by at least an order of magnitude and allows performance to scale. Matching pushes and pops enables the stack to perform even better on a mixed workload.
  }
  \label{fig:vector}
\end{figure}

@@

We also experimented with performing both forms of combining together; however, combining at the master did not affect performance significantly. This was somewhat surprising as combined updates still force others to wait. However, with both levels of combining enabled, relatively few requests are combined at the master--for the all-push workload, on the order of a thousand, and for the mixed workload on the stack, the average is less than 20.

@@

\paragraph{HashSet and HashMap}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/hash_perf.pdf}
  \caption{\emph{GlobalHashSet and GlobalHashMap.}
    Results are shown for a throughput workload inserting and looking up 256 million random keys in a particular range into a global hash with the same number of cells, with 2048 workers per core and 16 cores per node.
  }
  \label{fig:hash_perf}
\end{figure}

@@

\subsection{Application Kernel Performance}
The original goal was to have scalable global data structures in the Grappa library that could be used in applications. Generic implementations of data structures are often insufficient for scaling, so HPC applications often implement their own customized versions. However, this forces each application to carefully reason about the ways in which the structures will be accessed, and often relies on relaxing consistency.
The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
In these evaluations, we compare the flat-combining structures against tuned, relaxed-consistency versions.

\paragraph{Breadth-First Search}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/bfs_perf.pdf}
  \caption{\emph{BFS} on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm. Performance is measured in millions of Traversed Edges Per Second (MTEPS).}
  \label{fig:bfs_perf}
\end{figure}
The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark~\cite{graph500list}. This benchmark does a search starting from a random vertex in a synthetic graph and builds a search tree of parent vertices for each vertex traversed during the search. While this is a relatively simple problem, it exercises the random-access throughput of a system as well as being a primitive in many other graph algorithms. The BFS algorithm contains a global queue which represents the frontier of vertices to be visited in each level.
Our implementation employs the direction-optimizing algorithm by Beamer et al.~\cite{Beamer:Graph500} which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
The frontier queue in BFS is amenable to further optimization to take advantage of the fact that the algorithm does pushes and pops in separate phases, allowing consistency to be relaxed.
We compare our implementation of BFS using the flat-combined global queue described above with a highly tuned Grappa implementation that uses a custom asynchronous queue.

Figure~\ref{fig:bfs_perf} shows the results of scaling the BFS kernel up to 64 nodes. The simple queue implementation without flat combining is completely unscalable. However, the flat combining queue tracks the tuned version. This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible without sacrificing scaling.
% , while further optimizations can be applied incrementally.

\paragraph{Connected Components}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/cc_perf.pdf}
  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph. Performance is measured in MTEPS.}
  \label{fig:cc_perf}
\end{figure}
Connected Components (CC) is another simple graph analysis kernel that illustrates a different use of global data structures in irregular applications. We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively parallel MTA-2 machine. In the first phase, parallel traversals attempt to claim vertices and label them.
Whenever two traversals encounter each other, an edge between the two roots is inserted in a set. After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph. The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on this reduced graph, and the final phase propagates the component labels out to the full graph.
Creation of the reduced graph dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set. Therefore, we compare our straightforward implementation using the generic Set with and without flat combining against a tuned asynchronous implementation.

>>>>

\section{Evaluation}
To evaluate the impact flat combining has on the global data structures, we ran a series of experiments to test the raw performance of the data structures themselves under different workloads, and then measure their impact on performance of two graph benchmarks.
Experiments were run on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand network cards, connected via a QLogic InfiniBand switch.

@@

\paragraph{Queue and Stack.}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/vector_perf.pdf}
  \caption{\emph{Queue and Stack performance.}
    Results are shown on a log scale for a throughput workload performing 256 million operations with 2048 workers per core and 16 cores per node. Local flat combining improves throughput by at least an order of magnitude and allows performance to scale. Matching pushes and pops enables the stack to perform even better on a mixed workload.
  }
  \label{fig:vector}
\end{figure}

@@

We also experimented with performing both forms of combining together; however, combining at the master did not affect performance significantly. This was somewhat surprising as combined updates still force others to wait. However, with both levels of combining enabled, relatively few requests are combined at the master--for the stack, a maximum of 6\% per pass for the all-push workload and 0.1\% for the mixed workload.

@@

\paragraph{HashSet and HashMap.}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{data/plots/hash_perf.pdf}
  \caption{\emph{GlobalHashSet and GlobalHashMap.}
    Results are shown for a throughput workload inserting and looking up 256 million random keys in a particular range into a global hash with the same number of cells, with 2048 workers per core and 16 cores per node.
  }
  \label{fig:hash_perf}
\end{figure}

@@

\subsection{Application Kernel Performance}
% The original goal was to have scalable global data structures in the Grappa library that could be used in applications. Generic implementations of data structures are often insufficient for scaling, so HPC applications often implement their own customized versions. However, this forces each application to carefully reason about the ways in which the structures will be accessed, and often relies on relaxing consistency.
The Grappa data structures are synchronized to provide the most general use and match the expectations of programmers and algorithms.
In these evaluations, we compare the flat-combining structures against custom, tuned versions that leverage relaxed consistency needs of the applications.

\paragraph{Breadth-First Search.}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.38\textwidth]{data/plots/bfs_perf.pdf}
  \caption{\emph{BFS} on a Graph500-spec graph of scale 26 (64 million vertices, 1 billion edges), with the direction-optimizing BFS algorithm. Performance is measured in millions of Traversed Edges Per Second (MTEPS).}
  \label{fig:bfs_perf}
\end{figure}
The first application kernel is the Graph500 Breadth-First-Search (BFS) benchmark~\cite{graph500list}. This benchmark does a search starting from a random vertex in a synthetic graph and builds a search tree of parent vertices for each vertex traversed during the search.
% While this is a relatively simple problem, it exercises the random-access throughput of a system as well as being a primitive in many other graph algorithms.
The BFS algorithm contains a global queue which represents the frontier of vertices to be visited in each level.
Our implementation employs the direction-optimizing algorithm by Beamer et al.~\cite{Beamer:Graph500}.
 % which performs particularly well for the scale-free RMAT graphs generated by the benchmark.
The frontier queue is write-only in one phase and read-only in the next, making it amenable to relaxed consistency.
We compare performance of BFS using the Grappa FC queue described above with a highly tuned Grappa implementation that uses a custom asynchronous queue.

Figure~\ref{fig:bfs_perf} shows the results of scaling up to 64 nodes. Using the simple queue without flat combining is completely unscalable, but with FC, it tracks the tuned version. This illustrates that providing a safe, synchronized data structure for initially developing algorithms for PGAS is possible without sacrificing scaling.
% , while further optimizations can be applied incrementally.

\paragraph{Connected Components.}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.38\textwidth]{data/plots/cc_perf.pdf}
  \caption{\emph{Connected Components} on the same scale 26 Graph500 graph. Performance is measured in MTEPS.}
  \label{fig:cc_perf}
\end{figure}
Connected Components (CC) is another simple graph analysis kernel that illustrates a different use of global data structures in irregular applications. We implement the three-phase CC algorithm~\cite{mtgl} which was designed for the massively parallel MTA-2 machine. In the first phase, parallel traversals attempt to claim vertices and label them.
Whenever two traversals encounter each other, an edge between the two roots is inserted in a set. 
% After all edges have been traversed in this way, the set of edges forms a new, typically much smaller, graph.
The second phase performs the classical Shiloach-Vishkin parallel algorithm~\cite{shiloach1982n} on the reduced graph formed by phase one's edge set, and the final phase propagates the component labels out to the full graph.
Creation of the reduced edge set dominates the runtime of this algorithm, but includes many repeated inserts at the boundary between traversals, so is a prime target for the flat-combining Set.
As in BFS, further optimizations involving relaxation of consistency can be applied, in this case to the Set. Therefore, we compare our straightforward implementation using the generic Set with and without flat combining against a tuned asynchronous implementation.

### 2013-06-15 doc/pgas13/flat-combining.tex
rewriting to conserve space, tighten language

One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches. Having a single thread do all of the operations on the data structure allows it to keep everything in its cache.

@@

Reduced synchronization comes from delegating work to another core.
To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock. Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock. When many threads contend for the lock, all but one will fail and will have to retry. Each attempt forces an expensive memory fence and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets. Under flat combining, instead, threads add requests to the publication list. They each try to acquire the lock, and the one that succeeds becomes the combiner. Instead of retrying, the rest spin on their request waiting for it to be filled. The combiner walks the publication list, performs all of the requests, and when done, releases the lock. This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization--threads must synchronize to add to the publication list. However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost. This publication list mechanism can be re-used in different data structures, saving each one from needing to implement its own clever synchronization mechanism.

The above example of delegation is compelling in itself. However, the crux of prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.

@@

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}. The task must send a message to the master to acquire the lock. If successful, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All others block at the first message until the lock is released. Grappa's user-level threading allows requests to block without consuming compute resources. However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.

\paragraph{Centralized Combining.}
A first thought might be to directly apply the idea of flat combining to the serialization at the \emph{master}. The worker that acquires the lock on the master can walk through the requests of all the other workers waiting to acquire the global lock and perform combining on them. In the case of the GlobalStack, this would mean taking all of the push requests, matching them with the pops, and sending messages back to the remote workers with the results. Any leftover pushes or pops would be applied to the data structure, and another round of combining would begin. This approach reduces traffic to the data structure storage, but a single core must still process every request. This approach will not scale when every other core can generate requests just as fast as the master can combine them.

\paragraph{Distributed Combining.}
Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, each core can instead do its own combining first then synchronize with the master in bulk.
Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.

@@

In distributed flat-combining, each core builds its own publication list to track all the operations waiting to be committed to the global data structure.
In Grappa, for each global data structure, a local \emph{proxy} object is allocated from the core-local heap to intercept requests.
Conceptually, workers add their requests to a local publication list in the proxy, one is chosen to do the combining, and the rest block until their request is satisfied.
However, because Grappa's scheduler is non-preemptive, each worker has atomicity ``for free'' until it performs a blocking operation (such as communication).
This means that an explicit publication list with clever synchronization is unnecessary.
Instead, workers perform their own combining, merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed. Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops are matched on the proxy's local stack.

After all local combining has been done, one requesting worker on the core is chosen to commit the combined operation globally. In Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization. The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master. Combining in this way allows synchronization to be performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.

\paragraph{Memory Consistency Model.}
\label{sec:memory-model}
In the context of Grappa, sequential consistency guarantees that within a
task, operations will be in task order and that all tasks will observe the
same global order.

The Grappa memory model is essentially the same as the C++ memory
model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
consistency for data-race free programs). To be conservative, delegate
operations block the calling worker until they have become globally visible,
ensuring they can be used for synchronization. As such, delegate operations
within a task are guaranteed to be globally visible in program order and all
tasks observe the same global order.
Operations on synchronized data structures must provide the same guarantees.
Because it is not immediately obvious
that this distributed version of flat combining preserves sequential
consistency, we will now argue why it does.

To preserve linearizability for flat-combined operations, they must obey a consistent global order.
First of all, the execution of local combined operations must be serializable;
this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
% haven't discussed queue, set, or map yet...
%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure. The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking a batch of operations until all have completed.
For the \emph{stack}, matching pushes and pops locally is one exception to the rule that operations must block until globally committed. Because a pop ``destroys'' the corresponding push, these operations executed atomically together are independent of all others and can conceptually be placed anywhere in a total global ordering that preserves program order, so they can be safely matched locally.

Combining set and map operations expose more nuances in the consistency guarantees.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, providing the same guarantee as all other delegate operations. 
% This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.

In the same way as inserts, lookups can piggy-back on other lookups of the same key.
It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted, using the same intuition as for store/write buffers in modern processors.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then to preserve SC, that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches. Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.

Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use. If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates. This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).

These requirements only guarantee serializability for operations on a data structure in isolation. To guarantee sequential consistency with respect to accesses to all data structures, data-race freedom  must be guaranteed externally, as in the C++ memory model for shared memory.

\section{Grappa FC Framework}
In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.

The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine. As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
In this style, each global data structure must define the following:
\begin{enumerate}
  \item A local \emph{proxy} data structure for tracking updates to the global structure.
  \item \emph{Combining methods} that operate on the local proxy.
  \item A \emph{sync} method that globally commits the proxy's state.
\end{enumerate}
An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.

@@

\begin{figure}[t]
\centering
\begin{lstlisting}[style=grappa]
class GlobalStackProxy : Grappa::FCProxy {
  // pointer to data structure master
  GlobalAddress<GlobalStack> master;
  // local state for tracking pushes and pops
  T  pushed_values[1024];
  T* popped_results[1024];
  int npush, npop;

@@

With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake Workers, deliver results, manage when to synchronize, and choose the combiner.
The FC framework is responsible for ensuring that all of the combined operations eventually get committed. There are a number of ways progress could be guaranteed, but one of the simplest is to ensure that as long as there are any outstanding requests, at least one worker is committing a combined operation.
When that combined operation finishes, if there are still outstanding requests that have built up in the meantime, another blocked worker is chosen to do another combined synchronization.

One could imagine other synchronization policies that would ensure progress but allow more combining to occur before synchronizing. For instance, local combining could continue until all workers are blocked. However, this would require the flat-combining framework to track the activity of the workers on a core, and additionally would add latency unless the system saved enough active workers to tolerate the latency of the global commit action. As it is, the one-in-flight policy implemented performs well, and does not require any coupling with the Grappa scheduler. Exploration of other policies is left for future work.

While a combining operation is in flight, new requests are likely to continue to accumulate. The framework transparently wraps the proxy object so that when it starts a \texttt{sync} on one, it can direct requests to a fresh combiner.
This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
Requests cannot be combined with others that have been committed or are currently being committed; we refer
% Updates cannot be added to combiners that are in the process of being committed and must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them. 
 readers to Section~\ref{sec:memory-model} for why this must be so.

@@

The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first does the pushes, then does the pops.

\subsection{Global Set and Map}
The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a hash function. Our flat-combining version supports both \texttt{insert} and \texttt{lookup}. To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}). Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers. As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups. This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.

>>>>

% One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches. Having a single thread do all of the operations on the data structure allows it to keep everything in its cache.

@@

Simply by delegating work to another core, locality is improved and synchronization is reduced.
% Reduced synchronization comes from delegating work to another core.
To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock. Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock. When many threads contend for the lock, all but one will fail and have to retry. Each attempt forces an expensive memory fence and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets. Under flat combining, instead, threads add requests to a \emph{publication list}. They each try to acquire the lock, and the one that succeeds becomes the combiner. Instead of retrying, the rest spin on their request waiting for it to be filled. The combiner walks the publication list, performs all of the requests, and when done, releases the lock. This allows the one thread to keep the data structure in cache, reducing thrashing between threads. It also greatly reduces contention on the lock, but introduces a new point of synchronization--adding to the publication list. However, if a thread performs multiple operations, it can leave its publication record in the list and amortize the synchronization cost. This publication list mechanism can be re-used in other data structures, saving each from needing to implement its own clever synchronization.

The above example of delegation is compelling in itself. However, the crux of prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.

@@

All tasks accessing the stack hold a \texttt{GlobalAddress} to the master object, and invoke custom \emph{delegate methods} that, like the \texttt{read} delegate described earlier, block the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}. The task must send a message to the master to acquire the lock. If successful, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All others block at the first message until the lock is released. Grappa's user-level threading allows requests to block without consuming compute resources. However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.

\paragraph{Centralized Combining.}
A first thought might be to directly apply the idea of flat combining to the serialization at the \emph{master}. The worker that acquires the lock can walk through the requests of all the other workers waiting to acquire the lock and combine them. In the case of the Stack, this would mean matching pushes and pops, applying the remainder, and sending messages back to all remote workers with results, before starting another round of combining. This approach reduces traffic to the data structure storage, but a single core must still process every request, so it cannot scale if every other core is generating requests at the same rate.

\paragraph{Distributed Combining.}
Instead of all workers sending independent synchronization messages and putting the burden all on one core, each core can instead do its own combining first then synchronize with the master in bulk.
Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
% As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.

@@

In distributed flat-combining, each core builds its own publication list to track all the operations waiting to be committed to the global data structure.
In Grappa, for each global data structure, a local \emph{proxy} object is allocated from the core-local heap to intercept requests.
Conceptually, workers add their requests to a local publication list in the proxy, one is chosen to do the combining, and the rest block until their request is satisfied.
However, because Grappa's scheduler is non-preemptive, each worker has atomicity ``for free'' until it performs a blocking operation (such as communication).
This means that an explicit publication list with clever synchronization is unnecessary.
Instead, workers merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed. Figure~\ref{fig:stackfc} shows how pushes and pops are matched on the proxy's local stack.

After all local combining has been done, one requesting worker on the core is chosen to commit the combined operation globally. In Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization. The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master, but the granularity of global synchronization is coarser, reducing actual serialization.

\paragraph{Memory Consistency Model.}
\label{sec:memory-model}
In the context of Grappa, sequential consistency guarantees that within a
task, operations will be in task order and that all tasks will observe the
same global order.
The Grappa memory model is essentially the same as the C++ memory
model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
consistency for data-race free programs). To be conservative, delegate
operations block the calling worker until they have become globally visible,
ensuring they can be used for synchronization. As such, delegate operations
within a task are guaranteed to be globally visible in program order and all
tasks observe the same global order.
Operations on synchronized data structures must provide the same guarantees.
Because it is not immediately obvious
that this distributed version of flat combining preserves sequential
consistency, we now argue why it does.

To preserve linearizability for flat-combined operations, they must obey a consistent global order.
First of all, the execution of local combined operations must be serializable;
this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use. If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates.
% This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).
% haven't discussed queue, set, or map yet...
%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this serial ``concatenation'' of serialized batches to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure.
% Centralized combining preserves this by only allowing updates to proceed when they will not conflict and blocking a batch of operations until all have completed.
Matching pushes and pops locally for the Stack is one exception to the rule that operations must block until globally committed. Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a global order. It is trivial to respect program order with this placement, so they can be safely matched locally.

Combining set and map operations expose more nuances in the consistency guarantees.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
Note that if an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done to ensure that a subsequent operation it performs cannot be reordered with it, preserving program order.
% providing the same guarantee as all other delegate operations. 
% This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.

Similarly, lookups can piggy-back on other lookups of the same key.
Using intuition from store/write buffers in modern processors, it is tempting to allow a lookup to be satisfied locally by inspecting outstanding inserts.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then to preserve SC, that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches. Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.

These requirements only guarantee linearizability of operations on a single data structure. To guarantee sequential consistency with respect to all accesses, data-race freedom must be guaranteed by the program, as in the C++ memory model for shared memory.

\section{Grappa FC Framework}
To leverage the flat-combining paradigm in Grappa, we implemented a generic framework to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.

As mentioned before, the Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine.
% As mentioned before, the local \emph{publication list} is trivial to access atomically in Grappa's cooperative scheduling.
% Instead, in the Grappa FC Framework, 
A \emph{proxy} object is instantiated ahead of time on each core and each worker merges its request into that structure before blocking or becoming the combiner.
Each global data structure must define a \emph{proxy} with the following:
\begin{enumerate}
  \item A local \emph{proxy} data structure for tracking combined requests.
  \item \emph{Combining methods} that operate on the local proxy.
  \item A \emph{sync} method that globally commits the proxy's state.
\end{enumerate}
An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.

@@

\begin{figure}[t]
\centering
\begin{lstlisting}[style=grappa]
class GlobalStackProxy : Grappa::FCProxy {
  GlobalAddress<GlobalStack> master;
  // Local state for tracking requests
  T  pushed_values[1024];
  T* popped_results[1024];
  int npush, npop;

@@

With this data-structure-specific proxy object, the FC framework does the bookkeeping to block and wake workers, deliver results, manage when to synchronize, and choose the combiner.
The FC framework is responsible for ensuring that all of the combined operations eventually get committed. There are a number of ways progress could be guaranteed, but one of the simplest is to ensure that as long as there are any outstanding requests, at least one worker is committing a combined operation.
When that combined operation finishes, if there are still outstanding requests that have built up in the meantime, another blocked worker is chosen to do another combined synchronization.

% % left out for space
% One could imagine other synchronization policies that would ensure progress but allow more combining to occur before synchronizing. For instance, local combining could continue until all workers are blocked. However, this would require the flat-combining framework to track the activity of the workers on a core, and additionally would add latency unless the system saved enough active workers to tolerate the latency of the global commit action. As it is, the one-in-flight policy implemented performs well, and does not require any coupling with the Grappa scheduler. Exploration of other policies is left for future work.

While a combining operation is in flight, new requests continue to accumulate. The framework transparently wraps the proxy object so when one \texttt{sync} starts, it can direct requests to a fresh combiner.
This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
% Requests cannot be combined with others that have been committed or are currently being committed; we refer readers to Section~\ref{sec:memory-model} for why this must be so.
% Updates cannot be added to combiners that are in the process of being committed and must start out fresh each time; it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.

@@

The GlobalQueue has nearly the same implementation as the stack, but is unable to match locally, so \texttt{sync} first commits the pushes, then the pops.

\subsection{Global Set and Map}
The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a hash function. Our flat-combining version supports both \texttt{insert} and \texttt{lookup}. To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}). As with pops, lookups must provide pointers for results to be put, which is done with a hash map (again from the C++ standard library) from keys to lists of pointers. As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order. Instead, we only allow matching inserts with inserts and lookups with lookups, allowing \texttt{sync} to simply issue all inserts and lookups in parallel and block until all have completed.
